Prefetching SPARQL query cacher

### 2016-08-03 dissertation/cache-proxy.tex
Go through text flow #18
Go through text flow in cache #18

The query planner is written as completely stand-alone modules, based
on the facilities in the Attean framework, but without accessing
Attean internals. The system has in fact three query planners:
\pmodule{AtteanX::QueryPlanner::Cache},
\pmodule{AtteanX::QueryPlanner::Cache::LDF} and
\pmodule{AtteanX::Query::Cache::Analyzer::QueryPlanner}. Each is an
extension of the previous. The first is written to use a remote
endpoint in combination with a local cache. The second extends this to
also be able to generate and use TPF plans. The
final is a query planner for the analyser in Section~\ref{sec:analimpl}
that reruns the query planning with all triple patterns in a simulated
cache.

These query planners are a good example of the virtues of traits-based
programming, as they are typically customised along different axes:
access plans, cost estimation, planning of joins within Basic Graph
Patterns or left joins, rewriting (like gathering quad patterns to
blocks), etc. In the case of our query planner that combines a remote
endpoint with a local cache, named
\pmodule{AtteanX::QueryPlanner::Cache}, the basic query planner of the
Attean framework, \pmodule{Attean::QueryPlanner} is extended
(i.e. inherited from), but also composes the following roles:
\pmodule{Attean::API::NaiveJoinPlanner},
\pmodule{Attean::API::SimpleCostPlanner},
\pmodule{AtteanX::API::JoinRotatingPlanner},
\pmodule{AtteanX::Query::AccessPlan::SingleQuadBGP} and
\pmodule{AtteanX::Query::AccessPlan::Cache}, out of which the three
first are a part of the Attean framework, the next is described in
Appendix~\ref{sec:sparqlclient} and the last is a part of the
query cache system itself.

@@

The implementation of the join rotation detailed in
Section~\ref{sec:understandquery} is in the role
\pmodule{AtteanX::API::JoinRotatingPlanner}. In the planner
\pmodule{AtteanX::QueryPlanner::Cache} adjacent quad patterns are
coalesced into Basic Graph Pattern (BGP) plans, and BGP plans are
coalesced into larger BGP plans, so that they can be evaluated by the
remote SPARQL endpoint as a single unit. The extension
\pmodule{AtteanX::QueryPlanner::Cache::LDF} adds the option of
querying remote LDF servers.

@@

Throughout the system, the query planner takes the cost from the
models implementing the \pmodule{Attean::API::CostPlanner} role. A
key challenge is to balance the costs given by different models in the
system, and this is done in
\pmodule{AtteanX::Model::SPARQLCache::LDF}. Let us, however, start
from the bottom.

@@

Next, we detail the cost of a join where one or more children in the
tree is a remote SPARQL plan. It is usually not wanted to execute
several remote queries, it is assumed that the number of HTTP requests
should be small, and that plans where the joins have been rotated and
coalesced by \pmodule{AtteanX::API::JoinRotatingPlanner} should be
less expensive. Therefore, such plans should be penalised, but by how
much depends on the type plans. Conventionally, for a nested loop join
plan, the we first assign 
\begin{equation}\label{eq:cnl}
T_{A \bowtie_{nl} B} = 10 ((1 + C_A) C_B ) ,
\end{equation}
where $A$ and $B$ signifies the left and right subplans, and $C_A$ and
$C_B$ their respective cost estimates (which is originates from the
other heuristics in this section).  Following the default
implementation in Attean, we multiply the cost in Eq.~(\ref{eq:cnl})
by a factor of 10 if there are no shared variables, to mildly penalise
Cartesian joins.

@@

The system also depends on modules I have written that are not part of
the project: \pmodule{RDF::LinkedData}, \pmodule{RDF::Generator::Void}, \pmodule{Test::RDF},
\pmodule{URI::NamespaceMap} and \pmodule{RDF::NS::Curated}.

@@

The \pmodule{AtteanX::Store::SPARQL} module is a partial SPARQL
Protocol client that implements Attean APIs. The store implementation
itself composes the \mbox{TripleStore} API and implements methods to retrieve
the results of a single triple pattern and exact cardinality for a
given triple pattern, by using an aggregate query. In addition, it can
generate plans for Basic Graph Pattern algebra objects that have more
than one triple pattern. If that is the case, it will return an
instance of the class \pmodule{AtteanX::Plan::SPARQLBGP}, which is
also defined by the module. If it is not the case, the module also has
an implementation of the \pcode{access\_plans} method in a
\pmodule{AtteanX::Query::AccessPlan::SingleQuadBGP} role, that can be
composed by a query planner to provide a single triple pattern
\pmodule{AtteanX::Plan::SPARQLBGP} object. Finally, it also provides a
model implementation that composes the Model and \mbox{CostPlanner}
APIs. This contains a \pcode{cost\_for\_plan} implementation that will
provide a cost for \pmodule{AtteanX::Plan::SPARQLBGP} objects that are
proportional to the number of triple patterns in the plan, but
penalises plans that do not have triple patterns that are connected
through a variable (i.e. will cause a Cartesian join) with a factor
10. See Section~\ref{sec:costheuristics} for details on the cost
heuristics.

>>>>

The query planner is written as completely stand-alone modules, based
on the facilities in the Attean framework, but without accessing
Attean internals. The system has in fact three query planners:
\begin{itemize}
\item \pmodule{AtteanX::QueryPlanner::Cache},
\item \pmodule{AtteanX::QueryPlanner::Cache::LDF} and
\item \pmodule{AtteanX::Query::Cache::Analyzer::QueryPlanner}.
\end{itemize}
Each is an
extension of the previous. The first is written to use a remote
endpoint in combination with a local cache. The second extends this to
also be able to generate and use TPF plans. The
final is a query planner for the analyser in Section~\ref{sec:analimpl}
that reruns the query planning with all triple patterns in a simulated
cache.

These query planners are a good example of the virtues of traits-based
programming, as they are typically customised along different axes:
access plans, cost estimation, planning of joins within Basic Graph
Patterns or left joins, rewriting (like gathering quad patterns to
blocks), etc. In the case of our query planner that combines a remote
endpoint with a local cache, the module named above
\pmodule{AtteanX::QueryPlanner::Cache}, the basic query planner of the
Attean framework, \pmodule{Attean::QueryPlanner} is extended
(i.e. inherited from), but also composes the following roles:
\begin{itemize}
\item 
\pmodule{Attean::API::NaiveJoinPlanner},
\item \pmodule{Attean::API::SimpleCostPlanner},
\item \pmodule{AtteanX::API::JoinRotatingPlanner},
\item \pmodule{AtteanX::Query::AccessPlan::SingleQuadBGP} and
\item \pmodule{AtteanX::Query::AccessPlan::Cache},
\end{itemize}
out of which the three
first are a part of the Attean framework, the next is described in
Appendix~\ref{sec:sparqlclient} and the last is a part of the
query cache system itself.

@@

The implementation of the join rotation detailed in
Section~\ref{sec:understandquery} is in the role
\pmodule{AtteanX::API::JoinRotatingPlanner}. In the planner
\pmodule{AtteanX::Query\linebreak[4]Planner::Cache} adjacent quad patterns are
coalesced into Basic Graph Pattern (BGP) plans, and BGP plans are
coalesced into larger BGP plans, so that they can be evaluated by the
remote SPARQL endpoint as a single unit. The extension
\pmodule{AtteanX::QueryPlanner::Cache::LDF} adds the option of
querying remote LDF servers.

@@

Throughout the system, the query planner takes the cost from the
models implementing the \pmodule{Attean::API::CostPlanner} role. A
key challenge is to balance the costs given by different models in the
system, and this is done in the module
\pmodule{AtteanX::Model::SPARQLCache::LDF}. Let us, however, start
from the bottom.

@@

Next, we detail the cost of a join where one or more children in the
tree is a remote SPARQL plan. It is usually not wanted to execute
several remote queries, it is assumed that the number of HTTP requests
should be small, and that plans where the joins have been rotated and
coalesced by the implementation in the role \pmodule{AtteanX::API::JoinRotatingPlanner} should be
less expensive. Therefore, such plans should be penalised, but by how
much depends on the type plans. Conventionally, for a nested loop join
plan, the we first assign 
\begin{equation}\label{eq:cnl}
T_{A \bowtie_{nl} B} = 10 ((1 + C_A) C_B ) ,
\end{equation}
where $A$ and $B$ signifies the left and right subplans, and $C_A$ and
$C_B$ their respective cost estimates (which is originates from the
other heuristics in this section).  Following the default
implementation in Attean, we multiply the cost in Eq.~(\ref{eq:cnl})
by a factor of 10 if there are no shared variables, to mildly penalise
Cartesian joins.

@@

The system also depends on modules I have written that are not part of
the project, but that I have enhanced as time has allowed: 
\begin{itemize}
\item \pmodule{RDF::LinkedData},
\item \pmodule{RDF::Generator::Void},
\item \pmodule{Test::RDF},
\item \pmodule{URI::NamespaceMap} and
\item \pmodule{RDF::NS::Curated}.
\end{itemize}

@@

The \pmodule{AtteanX::Store::SPARQL} module is a partial SPARQL
Protocol client that implements Attean APIs. The store implementation
itself composes the supplied \mbox{TripleStore} API and implements methods to retrieve
the results of a single triple pattern and exact cardinality for a
given triple pattern, by using an aggregate query. In addition, it can
generate plans for Basic Graph Pattern algebra objects that have more
than one triple pattern. If that is the case, it will return an
instance of the class \pmodule{AtteanX::Plan::SPARQLBGP}, which is
also defined by the module. If it is not the case, the module also has
an implementation of the \pcode{access\_plans} method in a
\pmodule{AtteanX::Query::AccessPlan::SingleQuadBGP} role, that can be
composed by a query planner to provide a single triple pattern
\pmodule{AtteanX::Plan::SPARQLBGP} object. Finally, it also provides a
model implementation that composes the Model and \mbox{CostPlanner}
APIs. This contains a \pcode{cost\_for\_plan} implementation that will
provide a cost for \pmodule{AtteanX::Plan::SPARQLBGP} objects that are
proportional to the number of triple patterns in the plan, but
penalises plans that do not have triple patterns that are connected
through a variable (i.e. will cause a Cartesian join) with a factor
10. See Section~\ref{sec:costheuristics} for details on the cost
heuristics.
